---
title: "Introduction to fdth"
author: "Faria, J. C.; Allaman, I. B.; Jelihovschi, E. G."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to fdth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Overview

`fdth` builds **frequency distribution tables** (fdt) and their associated
graphics from vectors, data frames, and matrices for both **numerical** and
**categorical** variables.

Core functions:

| Function | Purpose |
|---|---|
| `fdt()` | Frequency table for numerical data |
| `fdt_cat()` | Frequency table for categorical data |
| `make.fdt()` | Reconstruct a table from frequencies alone |
| `make.fdt_cat()` | Reconstruct a categorical table from frequencies |
| `mfv()` | Most frequent value (mode) |
| `sd()` / `var()` | Standard deviation / variance for grouped data |

```{r load}
library(fdth)
```

---

## 1. Numerical data — `fdt()`

### 1.1 Basic usage

```{r fdt-basic}
set.seed(42)
x <- rnorm(200, mean = 10, sd = 2)

ft <- fdt(x)
ft
```

The default table has six columns:

| Column | Description |
|---|---|
| `Class limits` | Interval notation |
| `f` | Absolute frequency |
| `rf` | Relative frequency |
| `rf(%)` | Relative frequency (%) |
| `cf` | Cumulative frequency |
| `cf(%)` | Cumulative frequency (%) |

### 1.2 Choosing the number of classes

```{r fdt-breaks}
# Sturges (default)
fdt(x, breaks = "Sturges")

# Scott
fdt(x, breaks = "Scott")

# Freedman-Diaconis
fdt(x, breaks = "FD")

# Fixed number of classes
fdt(x, k = 8)
```

### 1.3 Custom interval boundaries

```{r fdt-custom}
# Fixed start, end and width
ft2 <- fdt(x,
           start = 4,
           end   = 16,
           h     = 2)
ft2
```

### 1.4 Formatting class limits

Use `format.classes = TRUE` together with `pattern` to control the number of
decimal places displayed in the class limits:

```{r fdt-format}
# Two decimal places
print(ft,
      format.classes = TRUE,
      pattern        = "%.2f")

# Summary with the same formatting
summary(ft,
        format.classes = TRUE,
        pattern        = "%.2f")
```

### 1.5 Right-closed intervals

By default intervals are left-closed `[a, b)`. Use `right = TRUE` for
right-closed `(a, b]`:

```{r fdt-right}
fdt(x, right = TRUE)
```

### 1.6 Missing values

```{r fdt-na}
x_na <- c(x, NA, NA)

# This errors by design:
tryCatch(fdt(x_na), error = function(e) message("Error: ", e$message))

# Remove NAs explicitly:
fdt(x_na, na.rm = TRUE)
```

---

## 2. Plots — `plot.fdt.default()`

All plot types are selected with the `type` argument.

### 2.1 Absolute frequency histogram and polygon

```{r plot-fh-fp, fig.show="hold", out.width="48%"}
plot(ft, type = "fh", main = "Frequency histogram")
plot(ft, type = "fp", main = "Frequency polygon")
```

### 2.2 Relative frequency (proportion and percentage)

```{r plot-rf, fig.show="hold", out.width="48%"}
plot(ft, type = "rfh",  main = "Relative frequency histogram")
plot(ft, type = "rfph", main = "Relative frequency (%) histogram")
```

### 2.3 Density

```{r plot-density}
plot(ft, type = "d", main = "Density histogram")
```

### 2.4 Cumulative frequency

```{r plot-cf, fig.show="hold", out.width="48%"}
plot(ft, type = "cfp",  main = "Cumulative frequency polygon")
plot(ft, type = "cfpp", main = "Cumulative frequency (%) polygon")
```

### 2.5 Value labels on bars

```{r plot-labels}
plot(ft,
     type    = "fh",
     v       = TRUE,
     v.round = 0,
     main    = "Histogram with counts")
```

---

## 3. Summary statistics from grouped data

Once an `fdt` object exists, the usual statistics can be computed directly
from the **grouped** (tabulated) data — no access to the original vector is
needed.

```{r stats}
ft3 <- fdt(x)

mean(ft3)
median(ft3)
mfv(ft3)          # mode(s)
var(ft3)
sd(ft3)

# Quartiles (default)
quantile(ft3)

# Deciles
quantile(ft3,
         i     = 1:9,
         probs = seq(0, 1, 0.1))
```

---

## 4. Multiple numerical variables — `fdt.data.frame()`

When the input is a **data frame** or **matrix**, `fdt()` builds one table
per numeric column and returns an `fdt.multiple` object.

### 4.1 All numeric columns

```{r fdt-df}
ft_iris <- fdt(iris[, 1:4])
ft_iris
```

### 4.2 Grouped by a factor

Use the `by` argument to stratify each numeric variable by a categorical
column:

```{r fdt-by}
ft_by <- fdt(iris[, c(1, 2, 5)],
             k  = 5,
             by = "Species")
ft_by
```

### 4.3 Plotting multiple tables

```{r plot-multiple, fig.width=8, fig.height=6}
plot(ft_iris, type = "fh")
```

### 4.4 Statistics on multiple tables

```{r stats-multiple}
mean(ft_iris)
```

---

## 5. Categorical data — `fdt_cat()`

### 5.1 Basic usage

```{r fdt-cat-basic}
set.seed(7)
fruits <- sample(c("apple", "banana", "cherry", "date", "apple",
                   "banana", "apple"),
                 150,
                 replace = TRUE)

ft_cat <- fdt_cat(fruits)
ft_cat
```

By default the table is sorted by descending frequency.

### 5.2 Preserving natural order

```{r fdt-cat-nosort}
fdt_cat(fruits, sort = FALSE)
```

### 5.3 Formatting

```{r fdt-cat-format}
print(ft_cat, round = 3)
```

### 5.4 Plots for categorical data

```{r plot-cat-bar}
plot(ft_cat, type = "fb", main = "Frequency bar chart")
```

```{r plot-cat-dotchart}
plot(ft_cat, type = "fd", main = "Frequency dotchart")
```

```{r plot-cat-pareto}
plot(ft_cat, type = "pa", main = "Pareto chart")
```

---

## 6. Reconstructing a table from frequencies

If the original data is no longer available but the frequency table is known,
`make.fdt()` and `make.fdt_cat()` rebuild complete `fdt` objects.

```{r make-fdt}
# Numerical
ft_ref <- fdt(x)

ft_new <- make.fdt(f     = ft_ref$table$f,
                   start = ft_ref$breaks["start"],
                   end   = ft_ref$breaks["end"])

print(ft_new, format.classes = TRUE, pattern = "%.2f")
```

```{r make-fdt-cat}
# Categorical
ft_new_cat <- make.fdt_cat(f          = ft_cat$f,
                            categories = ft_cat$Category)
ft_new_cat
```

---

## 7. LaTeX export

For publication-ready LaTeX tables use `xtable::xtable()` on any `fdt`
object. A dedicated vignette covers this workflow in detail:

```{r xtable-ref, eval=FALSE}
vignette("latex_fdt", package = "fdth")
```

---

## Session information

```{r session}
sessionInfo()
```
